---
title: ""
bibliography: "misc/ref.bib"
#csl: "misc/vancouver.csl"
output: 
  word_document:
    reference_docx: "misc/ref.docx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  include = FALSE
)
```

**Title**: [title]

**Authors**: Primer autor,^1^ Segundo autor,^2^ Tercer autor,^3^ Cuarto autor,^4^ Quinto autor,^5^ Sexto autor.^6^ \*

::: {custom-style="Afiliation"}
^1^ Afiliación, lugar.

^2^ Afiliación, lugar.

^3^ Afiliación, lugar.

^4^ Afiliación, lugar.

^5^ Afiliación, lugar.

^6^ Afiliación, lugar.
:::

### \*Autor de correspondencia 

Autor, Afiliación, lugar. e-mail: autor\@mail.com. 
Dirección: XXXX. Teléfono: XXXX.

\newpage

## Abstract 

**Objective**: [...]. **Material and methods**: [...]. **Results**: [...]. **Conclusion**: [...].

**Keywords**: [...].

\newpage

# Introduction

<!---- Antecedentes históricos ---->
[...].

<!---- Conceptualización del problema ---->
[...].

<!---- Hipótesis y objetivos ---->
[...].

# Material y methods

## Participants

[...]. 

## Instruments

### Instrument 1 

[...].

### Instrument 2

[...].


## Procedures 

[...].

[...].  

## Statistical analysis

Data is presented as median (*Mdn*) and interquartile range (*IQR*) for continuous variables, and as absolute and relative sample size for categorical/discrete variables.

A non-parametric approach was used since the underlying distribution of continuous outcomes, evaluated through analytical and graphical methods, did not follow a Gaussian distribution. 

In order assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate the goodness-of-fit ($\chi^2_{gof}$) and the independence of factors ($\chi^2_{Pearson}$). 

Generalized additive models (GAM) were used to describe linear and non-linear relationships between developmental characteristics through smooth terms that are represented using penalized regression splines [@wood2011fast]. The restricted maximum likelihood estimation method was used for the selection of the smoothing parameter, and the thin-plate regression spline as the smoothing basis method, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin]. GAMs were fitted using the mcgv package [@wood2017generalized]. To describe the smooth term, we use approximative derivative to summarise de trend in terms of linear segments.

[...], using the *R* programming language for statistical computing, [@rlanguage] and complementary R packages for model fitting, visualization and complementary computations [@wood2017generalized; @hadley2016ggplot2; @daniel2021see; @dominique2020estimation].

# Resultados

```{r load-packages}
source(file = "manuscript/misc/setup.R")
```

```{r}
# Sample
n_sample <- dataset[, .N]
#> 234

# Sample by sex
n_sex <- dataset[j = .(stat = paste0(.N, " (", round(.N*100/n_sample, 1), "%)")), 
                 keyby = .(sex = sexo_paciente)]
#>    sex        stat
#> 1:   F  94 (40.2%)
#> 2:   M 140 (59.8%)

# Prop test on sex
prop_sex <- dataset[, table(sexo_paciente)] |> 
  chisq.test() |> 
  report_chi(effsize = TRUE)
#> [1] "$\\chi^2_{gof}$ (1) = 9.04, *p* = 0.003, $C_{Pearson}$ = 0.19, CI~0.95%~[0.09, 1]"

```

From a total of `r n_sample` subjects with congenital hypotonia, `r n_sex["F", stat]` were females and `r n_sex["M", stat]` males (`r prop_sex`). The developmental characteristics of the sample can be seen in [Table 1](#tab1).

```{r}
mod <- gam(comunicacion_total ~ s(edad_cronologica_meses), 
    data = dataset,
    method = "REML")

# Overall slope
gam_coef <- report_slope(mod, trend = c("edad_cronologica_meses"))

# Linear segments
modelbased::estimate_smooth(
  data = modelbased::estimate_relation(mod, length = 100),
  x = "edad_cronologica_meses"
)

plot_gam(dataset, edad_cronologica_meses, comunicacion_total, round_to = 5,
         "Chronological age (months)", "Communication score", method = "REML")
```

When modelling the effect of chronological age on developmental skills, we found an averaged negative marginal effect on the communication score (`r gam_coef`), whereas from 1 to 5.7 moths old

# Discusion

[...].

[...].

[...].

# Conclusion

[...]. 

# Acknowledgment

[...].

# Conflictos de interés

[...].

# Referencias

<div id="refs"></div>
