---
title: "Main relationships"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, 
                      comment = "#>",
                      dpi = 300)
options(width = 140)
```

```{r setup}
library(ASQ3)
library(data.table)

# Round to nearest multiple
near <- function(i, j) {
  stopifnot("'i' must be a number" = is.numeric(i),
            "'j' must be a number" = is.numeric(j))
  
  i_hat = i / j
  output = round(i_hat) * j
  return(output)
}

# Plot GAM model of bivariate model
plot_gam <- function(data, x, y, xlab, ylab, round_to = 1, ...) {
  
  if (missing(xlab)) xlab <- deparse(substitute(x))
  if (missing(ylab)) ylab <- deparse(substitute(y))
  
  scaled = copy(data)
  if (round_to > 1) {
    j_exec = substitute(near(x, round_to))
    x_var = deparse(substitute(x))
    scaled[, (x_var) := eval(j_exec)]
  }
  
  ggplot2::ggplot(data, ggplot2::aes({{x}}, {{y}})) +
    ggplot2::geom_count(alpha = 0.1) +
    ggplot2::stat_summary(data = scaled, fun.data = ggplot2::mean_se) +
    ggplot2::geom_smooth(method = "gam", 
                formula = y ~ s(x),
                col = "red4",
                method.args = list(...)) +
    ggplot2::labs(x = xlab, y = ylab, caption = paste0("Mean ± SE in ", round_to, "-unit intervals")) +
    see::theme_lucid()
}
```

## Working with first assessment

For this page, we'll be fitting different models using generalized additive models (GAM's) using the `mgcv` package. The smoothing parameter will be estimated using restricted maximum likelihood algorithm (i.e., REML). 

### Communication score & chronological age (in months)

Model: comunicacion_total ~ edad_cronologica_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_cronologica_meses <= 60], 
         x = edad_cronologica_meses, 
         y = comunicacion_total, 
         ylab = "Communication", 
         xlab = "Months old",
         round_to = 5,
         method = "REML")
  
library(mgcv)
mgcv::gam(formula = comunicacion_total ~ s(edad_cronologica_meses),
          data = dataset[n_evaluacion == 1 & edad_cronologica_meses <= 60],
          method = "REML") |> 
  summary()
```

### Problem solving score & chronological age (in months)

Model: resolucion_problemas_total ~ edad_cronologica_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_cronologica_meses <= 60], 
         x = edad_cronologica_meses, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving", 
         xlab = "Months old",
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(edad_cronologica_meses),
          data = dataset[n_evaluacion == 1 & edad_cronologica_meses <= 60],
          method = "REML") |> 
  summary()
```

### Communication score & prematurity (in weeks)

Model: comunicacion_total ~ semanas_prematurez

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = semanas_prematurez, 
         y = comunicacion_total, 
         ylab = "Communication",
         xlab = "Prematurity (in weeks)", 
         round_to = 2,
         method = "REML")
  
mgcv::gam(formula = comunicacion_total ~ s(semanas_prematurez),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Communication score & corrected age (in months) 

Model: comunicacion_total ~ edad_corregida_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60], 
         x = edad_corregida_meses, 
         y = comunicacion_total, 
         ylab = "Communication",
         xlab = "Corrected age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = comunicacion_total ~ s(edad_corregida_meses),
          data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60],
          method = "REML") |> 
  summary()
```


### Fine motor skills score & corrected age (in months) 

Model: motora_fina_total ~ edad_corregida_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60], 
         x = edad_corregida_meses, 
         y = motora_fina_total, 
         ylab = "Fine motor skills",
         xlab = "Corrected age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = motora_fina_total ~ s(edad_corregida_meses),
          data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60],
          method = "REML") |> 
  summary()
```


### Problem solving score & corrected age (in months) 

Model: resolucion_problemas_total ~ edad_corregida_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60], 
         x = edad_corregida_meses, 
         y = resolucion_problemas_total, 
         ylab = "Fine motor skills",
         xlab = "Corrected age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(edad_corregida_meses),
          data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60],
          method = "REML") |> 
  summary()
```

### Socio-individual skills score & corrected age (in months) 

Model: socio_individual_total ~ edad_corregida_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60], 
         x = edad_corregida_meses, 
         y = socio_individual_total, 
         ylab = "Socio-individual skills",
         xlab = "Corrected age (months old)", 
         round_to = 4,
         method = "REML")
  
mgcv::gam(formula = socio_individual_total ~ s(edad_corregida_meses),
          data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60],
          method = "REML") |> 
  summary()
```

### Communication score & developmental ASQ-3 age (in months) 

Model: comunicacion_total ~ asq3_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = asq3_meses, 
         y = comunicacion_total, 
         ylab = "Communication",
         xlab = "developmental ASQ-3 age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = comunicacion_total ~ s(asq3_meses),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```


### Problem solving score & developmental ASQ-3 age (in months) 

Model: resolucion_problemas_total ~ asq3_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = asq3_meses, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "developmental ASQ-3 age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(asq3_meses),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Fine motor skills score & developmental ASQ-3 age (in months) 

Model: motora_fina_total ~ asq3_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = asq3_meses, 
         y = motora_fina_total, 
         ylab = "Fine motor skills",
         xlab = "developmental ASQ-3 age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = motora_fina_total ~ s(asq3_meses),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Fine motor skills score & Communication score 

Model: motora_fina_total ~ comunicacion_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = comunicacion_total, 
         y = motora_fina_total, 
         ylab = "Fine motor skills",
         xlab = "Communication", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = motora_fina_total ~ s(comunicacion_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Socio-individual skills score & Communication score 

Model: socio_individual_total ~ comunicacion_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = comunicacion_total, 
         y = socio_individual_total, 
         ylab = "Socio-individual skills",
         xlab = "Communication", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = socio_individual_total ~ s(comunicacion_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Gross motor skills score score & Communication score 

Model: motora_gruesa_total ~ comunicacion_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = comunicacion_total, 
         y = motora_gruesa_total, 
         ylab = "Gross motor skills",
         xlab = "Communication", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = motora_gruesa_total ~ s(comunicacion_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

Looks more like a linear relationship than a non-linear one. This is notorious when we look at the  effective degrees of freedom (edf), i.e. the closer to 1 more alike to a linear regression.

### Problem solving score & Fine motor skills score 

Model: resolucion_problemas_total ~ motora_fina_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = motora_fina_total, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "Fine motor skills", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(motora_fina_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Socio-individual skills score & Fine motor skills score 

Model: socio_individual_total ~ motora_fina_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = motora_fina_total, 
         y = socio_individual_total, 
         ylab = "Socio-individual skills",
         xlab = "Fine motor skills", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = socio_individual_total ~ s(motora_fina_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Problem solving score & Communication score 

Model: resolucion_problemas_total ~ comunicacion_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = comunicacion_total, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "Communication", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(comunicacion_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Problem solving score & Gross motor skills score

Model: resolucion_problemas_total ~ motora_gruesa_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = motora_gruesa_total, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "Gross motor skills", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(motora_gruesa_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Problem solving score & Socio-individual skills score

Model: resolucion_problemas_total ~ socio_individual_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = socio_individual_total, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "Socio-individual skills", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(socio_individual_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```
