---
title: "Main relationships"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, 
                      comment = "#>",
                      dpi = 300)
options(width = 140)
```

<details>
<summary>Setup details</summary>
```{r setup}
library(ASQ3)
library(data.table)

# Round to nearest multiple
near <- function(i, j) {
  stopifnot("'i' must be a number" = is.numeric(i),
            "'j' must be a number" = is.numeric(j))
  
  i_hat = i / j
  output = round(i_hat) * j
  return(output)
}

# Plot GAM model of bivariate model
plot_gam <- function(data, x, y, xlab, ylab, round_to = 1, ...) {
  
  if (missing(xlab)) xlab <- deparse(substitute(x))
  if (missing(ylab)) ylab <- deparse(substitute(y))
  
  scaled = copy(data)
  if (round_to > 1) {
    j_exec = substitute(near(x, round_to))
    x_var = deparse(substitute(x))
    scaled[, (x_var) := eval(j_exec)]
  }
  
  ggplot2::ggplot(data, ggplot2::aes({{x}}, {{y}})) +
    ggplot2::geom_count(alpha = 0.1) +
    ggplot2::stat_summary(data = scaled, fun.data = ggplot2::mean_se) +
    ggplot2::geom_smooth(method = "gam", 
                formula = y ~ s(x),
                col = "red4",
                method.args = list(...)) +
    ggplot2::labs(x = xlab, y = ylab, caption = paste0("Mean ± SE in ", round_to, "-unit intervals")) +
    see::theme_lucid()
}
```
</details>

## Working with first assessment

For this page, we'll be fitting different models using generalized additive models (GAM's) using the `mgcv` package for modeling non-linear relationships, even though it can be used to model linear relationships too. The smoothing parameter (the one that gets to decide how wiggly will be the curve) will be estimated using restricted maximum likelihood (i.e., REML). 

### About interpretation

The interpretation of the summary of GAM models in R can be seen in [this short video](https://campus.datacamp.com/courses/nonlinear-modeling-in-r-with-gams/interpreting-and-visualizing-gams?ex=1) (~ 4 minutes) from _datacamp_. 
However, let's get a short intro to a fast interpretation!

```
Approximate significance of smooth terms:
                            edf Ref.df     F p-value    
s(edad_cronologica_meses) 6.475  7.561 28.98  <2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

This small part of a summary output shows that the influence of `edad_cronologica_meses` is significant at a p-value of <2e-16. The `edf` stands for "Effective Degrees of Freedom" and reflects how complex (or wiggly) the relationship is. Thus, an `edf` of 1 reflects a linear relationship, while and `edf` of 2 reflects a cuadratic relationship, an `edf` of 3 cubic relationship and so on.

More information about GAMs can be seen here:

- Generalized Additive Models, Michael Clark: [website](https://m-clark.github.io/generalized-additive-models/introduction.html).
- Nonlinear Models in R: The Wonderful World of mgcv, Noam Ross: [youtube](https://youtu.be/q4_t8jXcQgc).

## Modeling
 
### Communication score & chronological age (in months)

Model: comunicacion_total ~ edad_cronologica_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_cronologica_meses <= 60], 
         x = edad_cronologica_meses, 
         y = comunicacion_total, 
         ylab = "Communication", 
         xlab = "Months old",
         round_to = 5,
         method = "REML")
  
library(mgcv)
mgcv::gam(formula = comunicacion_total ~ s(edad_cronologica_meses),
          data = dataset[n_evaluacion == 1 & edad_cronologica_meses <= 60],
          method = "REML") |> 
  summary()
```

### Problem solving score & chronological age (in months)

Model: resolucion_problemas_total ~ edad_cronologica_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_cronologica_meses <= 60], 
         x = edad_cronologica_meses, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving", 
         xlab = "Months old",
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(edad_cronologica_meses),
          data = dataset[n_evaluacion == 1 & edad_cronologica_meses <= 60],
          method = "REML") |> 
  summary()
```

### Communication score & prematurity (in weeks)

Model: comunicacion_total ~ semanas_prematurez

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = semanas_prematurez, 
         y = comunicacion_total, 
         ylab = "Communication",
         xlab = "Prematurity (in weeks)", 
         round_to = 2,
         method = "REML")
  
mgcv::gam(formula = comunicacion_total ~ s(semanas_prematurez),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Communication score & corrected age (in months) 

Model: comunicacion_total ~ edad_corregida_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60], 
         x = edad_corregida_meses, 
         y = comunicacion_total, 
         ylab = "Communication",
         xlab = "Corrected age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = comunicacion_total ~ s(edad_corregida_meses),
          data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60],
          method = "REML") |> 
  summary()
```


### Fine motor skills score & corrected age (in months) 

Model: motora_fina_total ~ edad_corregida_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60], 
         x = edad_corregida_meses, 
         y = motora_fina_total, 
         ylab = "Fine motor skills",
         xlab = "Corrected age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = motora_fina_total ~ s(edad_corregida_meses),
          data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60],
          method = "REML") |> 
  summary()
```


### Problem solving score & corrected age (in months) 

Model: resolucion_problemas_total ~ edad_corregida_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60], 
         x = edad_corregida_meses, 
         y = resolucion_problemas_total, 
         ylab = "Fine motor skills",
         xlab = "Corrected age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(edad_corregida_meses),
          data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60],
          method = "REML") |> 
  summary()
```

### Socio-individual skills score & corrected age (in months) 

Model: socio_individual_total ~ edad_corregida_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60], 
         x = edad_corregida_meses, 
         y = socio_individual_total, 
         ylab = "Socio-individual skills",
         xlab = "Corrected age (months old)", 
         round_to = 4,
         method = "REML")
  
mgcv::gam(formula = socio_individual_total ~ s(edad_corregida_meses),
          data = dataset[n_evaluacion == 1 & edad_corregida_meses <= 60],
          method = "REML") |> 
  summary()
```

### Communication score & developmental ASQ-3 age (in months) 

Model: comunicacion_total ~ asq3_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = asq3_meses, 
         y = comunicacion_total, 
         ylab = "Communication",
         xlab = "developmental ASQ-3 age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = comunicacion_total ~ s(asq3_meses),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```


### Problem solving score & developmental ASQ-3 age (in months) 

Model: resolucion_problemas_total ~ asq3_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = asq3_meses, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "developmental ASQ-3 age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(asq3_meses),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Fine motor skills score & developmental ASQ-3 age (in months) 

Model: motora_fina_total ~ asq3_meses

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = asq3_meses, 
         y = motora_fina_total, 
         ylab = "Fine motor skills",
         xlab = "developmental ASQ-3 age (months old)", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = motora_fina_total ~ s(asq3_meses),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Fine motor skills score & Communication score 

Model: motora_fina_total ~ comunicacion_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = comunicacion_total, 
         y = motora_fina_total, 
         ylab = "Fine motor skills",
         xlab = "Communication", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = motora_fina_total ~ s(comunicacion_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Socio-individual skills score & Communication score 

Model: socio_individual_total ~ comunicacion_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = comunicacion_total, 
         y = socio_individual_total, 
         ylab = "Socio-individual skills",
         xlab = "Communication", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = socio_individual_total ~ s(comunicacion_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Gross motor skills score score & Communication score 

Model: motora_gruesa_total ~ comunicacion_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = comunicacion_total, 
         y = motora_gruesa_total, 
         ylab = "Gross motor skills",
         xlab = "Communication", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = motora_gruesa_total ~ s(comunicacion_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

Looks more like a linear relationship than a non-linear one. This is notorious when we look at the  effective degrees of freedom (edf), i.e. the closer to 1 more alike to a linear regression.

### Problem solving score & Fine motor skills score 

Model: resolucion_problemas_total ~ motora_fina_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = motora_fina_total, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "Fine motor skills", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(motora_fina_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Socio-individual skills score & Fine motor skills score 

Model: socio_individual_total ~ motora_fina_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = motora_fina_total, 
         y = socio_individual_total, 
         ylab = "Socio-individual skills",
         xlab = "Fine motor skills", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = socio_individual_total ~ s(motora_fina_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Problem solving score & Communication score 

Model: resolucion_problemas_total ~ comunicacion_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = comunicacion_total, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "Communication", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(comunicacion_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Problem solving score & Gross motor skills score

Model: resolucion_problemas_total ~ motora_gruesa_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = motora_gruesa_total, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "Gross motor skills", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(motora_gruesa_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```

### Problem solving score & Socio-individual skills score

Model: resolucion_problemas_total ~ socio_individual_total

```{r}
plot_gam(data = dataset[n_evaluacion == 1], 
         x = socio_individual_total, 
         y = resolucion_problemas_total, 
         ylab = "Problem solving",
         xlab = "Socio-individual skills", 
         round_to = 5,
         method = "REML")
  
mgcv::gam(formula = resolucion_problemas_total ~ s(socio_individual_total),
          data = dataset[n_evaluacion == 1],
          method = "REML") |> 
  summary()
```
